---
output: html_document
editor_options: 
  chunk_output_type: console
---
## Data Preparation

The original data set consisted of tick data of **52** currency pairs in three months
( observations) between November, 2017 and March, 2018. There are 3 variables for each of them:

* Timestamp 
* Bid price 
* Ask price
* Mid price



The ticks were left untrasnformed. The purpose was to test the method in a real-life environment. Hence, e.g. aggregating would ruin the initial idea.

Bid and ask prices are taken from Bloomberg's FXGo platform using R Bloomberg API (RBlpapi) for 1 mio of base currency.
Bloomberg covers prices from hundreds of banks and for most currency pairs in the world. Some of them are crossed, with use of Euro or US Dollar, for instance **EUR/JPY** is the combination of:

$$EUR/JPY = EUR/USD\timesUSD/JPY$$

Sometimes pairs can effectively consist of 3 parts (legs), for instance **PLN/MXN**:
$$PLN/MXN = (USD/MXN\timesEUR/USD)/(EUR/PLN)$$

Using such pairs is usually problematic and reduces reliability in backtesting. Hence, for the purpose of the work only the most liquid pairs, based on G10 currencies, had been used:

* USD - United States dollar
* EUR - Euro
* JPY - Japanese yen
* GBP - Pound sterling
* CHF - Swiss franc
* AUD - Australian dollar
* NZD - New Zealand dollar
* CAD - Canadian dollar
* SEK - Swedish krona
* NOK - Norwegian Krone

In eFX the crucial element is spread, calculated as the difference between bid and ask prices. It depends on several factors, such as time of the day, one-off events, volatility, ability of liquidity providers to warehouse risk, or market sentiment. The data source had been selected so that it captured spread and reflected FX market as good as possible.

Below is the glimpse of the data used in the research:


```{r read_currencies}
library(dplyr)
library(tibble)
library(printr)
dataset <- tryCatch({readRDS('./../../../data/eurusd.RDS')}, error= function(x) getwd())
dataset <- as_tibble(dataset)

```

```{r dataset2}
dataset %>% head(10) %>% 
  arrange(Index) %>% 
  mutate(spread    = ask - bid,
         time_diff = c(NA, diff.POSIXt(Index)))


```


Cols: average spread (in us dollars), min spread, max spread, average number of ticks, mid price movements 

Rows: time of the day, sum

In FX market participants usually get quotations for different levels (tiers). The assumption of the work was to use the smallest one to reduce possible market impact.

## Code and the Research Process

The implementation of trading agents was based on R (both base and external libraries).

The graphical presentation was prepared with the use of ggplot2 library. 


The R-project consists of:

* frun.R, 
* get_data.R, 
* cointegration.R, 
* attributes.R, 
* discretization.R,
* functions.R, 
* state_space.R, 
* mcc.R, 
* q_learning.R



The main part was based on run.R script. Running it merges all above-mentioned scripts and executes the whole experiment. 
