<!--!Rnw root = ../../masters\\_thesis.Rnw -->


## FX Market Organization

Explaining the institutional structure of FX market requires introducing formal definitions of market organization. According to Lyons @Lyons2002, these are:

* Auction market - a participant can place a market and a limit order. The first action is aimed at buying X units at the best price. Alternatively, limit orders set a threshold, i.e. they are executed only if the market quotes reach a certain price. Limit orders are aggregated into an order book
* Single dealer market - in this kind of market organization, there is just one dealer. It is obliged to quote an asset, i.e. to match demand and supply. Its quotations are always the best bid and the best ask. The main task is to manage the risk to make profit off his spread.
* Multiple dealer market - it is extension of single dealer market. There is more than one dealer and they compete against each other. It might be centralized or decentralized. In the first version, all dealers are put into the same location while in the second it is not the case. When the market is decentralized, it is possible for price takers to gain profits by arbitrage transactions.

The FX market is a kind of decentralized multiple-dealer market. There is no single indicator that would show the best bid and the best ask. Hence, the market transparency is low. It is especially important at tail events. It is hard to determine when the market was at a given time and findings are usually spurious.

The foreign exchange market is perceived as the largest and most liquid one, with a year-on-year turnover of \texteuro 69 trillion.


```{r market_share2, include=FALSE,eval=TRUE, echo=FALSE, cache=TRUE}
market_share = c(
  0.1611,
  0.1454,
  0.0811,
  0.0765,
  0.0730,
  0.0622,
  0.0540,
  0.0365,
  0.0340,
  0.0338,
  0.0243,
  0.0240,
  0.0197,
  0.0166,
  0.0155
  )
```
The FX market is an over the counter, global (OTC) market, i.e. participants can trade currencies with relatively low level of legal obstacles. The market core is built up by the biggest banks in the world. Hence, the FX organization is often referred as
an inter-bank market.
The participants of the FX market differ by access, spreads, impact, turnover they generate, order size, and purpose. They can be divided into five main groups:

* Central banks - they control money supply, interest and reserve rates and hence can have the strongest market impact. Through their set of tools, they strengthen or weaken local currency. In the developed markets, their turnover is rather small due to the fact that intervenings happen rarely. On the other hand, order size is usually bigger than for other four groups due to the effect they want to achieve.
* Commercial banks - most of the flow in the market belongs to commercial banks. Although the environment in which FX trading occurs is highly dispersed in terms of location, over `r floor(sum(market_share)*100)`% of flow is generated by top 15 banks, as seen in \ref{table_turnover_banks}. It can be observed even for currencies that the banks do not have real interest in. It means that in fact banks stay with flat position.
Commercial banks make money on effective risk management. Essentially, it means taking flows from clients (retail/institutional) and managing risk books.
Over the years, the market have changed dramatically. Even though turnovers are higher than ten years, market practitioners tend to claim that liquidity has worsen. It is mostly due to the fact that new regulation, internal and external have been introduced.
Banks are required to stay with rather small positions, especially in non-G10 currencies. Their approach to risk is much more conservative than it used to be.
* Non-bank financial institutions - their significance as market participants is on the rise. Even though, non-bank financial institutions category is very broad and entities in it are very heterogenous, the most impactful are sophisticated hedge funds focused on effective market making (such as XTX Markets).
* Commercial companies - as price takers they are significantly worse than commercial banks due to the fact they trade bigger size and mainly hedge their main business.
* Retail traders - their main purpose is to speculate. The conditions they receive from financial institutions are generally worse.


```{r table_turnover, results='asis', echo=FALSE, cache.vars=T}
# library(xtable)
# require(pander)

df <- data.frame(Rank=seq_len(15),
                 Bank=c('Citi','Deutsche Bank','Barclays','JPMorgan','UBS','Bank of America Merrill Lynch',
                        'HSBC','BNP Paribas','Goldman Sachs','RBS','Societe Generale','Standard Chartered',
                        'Morgan Stanley','Credit Suisse','State Street'),
                 MarketShare=market_share)
strCaption <- paste0("Market share of top financial institutions in FX trading in 2014")

as_tibble(df %>% purrrlyr::dmap_at('MarketShare', ~paste0(.*100, '%'))) %>%
  set_colnames(c('Rank', 'Bank', 'Market Share'))
# print(xtable(df, digits = 2, caption = strCaption, label = "table_turnover_banks",
#       size="footnotesize", #Change size; useful for bigger tables
#       include.rownames  = FALSE, #Don't print rownames
#       include.colnames  = FALSE, #We create them ourselves
#       caption.placement = "top"
#       ), type = 'html')
#pandoc.table(df,style='grid')
```

In the last years, there have been observed shifting towards eFX (electronic trading of FX). Commercial banks, as mentioned in the previous subsection,
are subject to new regulations. Therefore, right now they are more concerned about increasing their turnover
than benefiting off speculation, e.g. trading based on macro research. eFX helps in this goal. It requires more technology while a number of traditional dealers
is effectively reduced. The activity require quantitative analysts, "quants", who can manage pricing engines in order to maximize profit while staying within risk constraints.
Over the last 4 years, eFX gained 13 percent point and in 2015 for the first time surpassed voice trading, with 53.2\% of client flow share  @JeffPatterson2015 @Chung2015.
