---
title: Reinforcement Learning Portfolio Optimization for FX Trading
author: Krzysztof Wojdalski
date: "`r Sys.Date()`"
# documentclass: pracamgrwne
output:
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true 
  # html_document:
    # toc: true
    # theme: united
    # highlight: tango
    # keep_md: false
bibliography: library.bib
header-includes:
  - \usepackage[utf8]{inputenc}
  - \usepackage{amsthm}
  - \usepackage{rotating}
  - \pagenumbering{gobble}
editor_options:
  chunk_output_type: console
  latex_engine: xelatex
---

\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newpage

# Abstract

The work is about reinforcement learning application in trading on the FX market.
The author starts with describing the FX market, analyzing market organization, participants, and changes in the last
years. He tries to explain current trends and the possible directions. 
The next part consists of theoretical pattern for the research - description of financial models, 
and the AI algorithms. 
Implementation of the RL-based approach in the third chapter, based on Q-learning, gives spurious results. 





```{r rr, include = FALSE, cache = FALSE, echo = FALSE, eval = TRUE}

if (!'pacman' %in%installed.packages()) install.packages('pacman')
library(pacman)
p_load(RefManageR, bibtex, knitr)
# bib <- ReadBib('./working_papers/library.bib')
# BibOptions(bib.style = "numeric",check.entries=FALSE)
opts_chunk$set(
  cache   = FALSE,
  warning = FALSE,
  error   = FALSE,
  message = FALSE,
  eval    = TRUE,
  echo    = FALSE,
  include = TRUE
)

# opts_knit$set(root.dir = NULL)
inputs <- './working_papers'
# Sys.setenv(TEXINPUTS=inputs,
#            BIBINPUTS=inputs,
#            BSTINPUTS=inputs)
# #Sys.getenv("RSTUDIO_PANDOC")
Sys.setenv(RSTUDIO_PDFLATEX = Sys.which("latexmk"))
#options(encoding="UTF-8")
#Sys.setlocale(locale='Polish_Poland.1250')
#Sys.getlocale('LC_ALL',')
```


```{r intro, child='chapters/subchapters/0_Introduction.Rmd'}
```

```{r fx_market_organization, child='chapters/subchapters/1_2_FX_Market_Organization.Rmd'}
```



```{r modern_portfolio_theory_and_capm,child='chapters/subchapters/2_1_1_Modern_Portfolio_Theory_and_CAPM.Rmd'}
```


```{r relevant_financial_indicators2, child = 'chapters/subchapters/2_2_Relevant_Financial_Indicators.Rmd'}
```


```{r machine_learning,child='chapters/subchapters/3_1_Machine_Learning.Rmd'}
```

```{r reinforcement_learning,child='chapters/subchapters/3_2_reinforcement_learning.Rmd'}
```


```{r objective_of_the_research,child='chapters/subchapters/4_1_Objective_of_the_Research.Rmd'}
```

```{r design_of_the_research,child='chapters/subchapters/4_2_design_of_the_research.Rmd'}
```

```{r dataset,child='chapters/subchapters/4_3_dataset.Rmd'}
```


```{r bibliography, child='chapters/subchapters/6_bibliography.Rmd'}
```






